---
title: Analysis of the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Database
author: "Diane Clow"
date: "June 14, 2015"
output: html_document
---
Reproducible Research


### Introduction
This analysis was produced for the second project in the Reproducible Research class offered on Coursera.  In this project, the Storm Database maintained by NOAA was analysed looking for information on: 1) What events are most harmful to humans, and 2) What events have the greated economical impact.

### Importing Data

The data is from the NOAA Storm Database and contains data from 1950 to November 2011.  The data has 902297 observations and 37 variables.

```{r}
StormData <- read.csv("repdata-data-StormData.csv.bz2")
library(ggplot2)
```

The column names in the data set are listed below

```{r}
colnames(StormData)
```

# Data Processing

### Question 1

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

To answer this question, first identify which variables will be used.  Reading the different column names, "FATALITIES" and "INJURIES",  Seemed the most usefuel.  Using the aggregate function, the total injuries and fatalities were calculated for each event type.

```{r}
fatalites <- aggregate(FATALITIES ~ EVTYPE, StormData, sum)
injuries <- aggregate(INJURIES ~ EVTYPE, StormData, sum)
```

Then the totals were reduced to the top 5 for both fatalites and injuries.

```{r}
fatal <- subset(fatalites, fatalites$FATALITIES > 600)
injury <- subset(injuries, injuries$INJURIES > 5000)
```

These two are combined, only keeping the values that are present in both fatalities and injuries.  At this point there are 3 events that were both in the top 5 for both fatalities and injuries.  harm2 is the same as harm, but the data frame has been sorted by the number of Fatalities, with the largest value in the first row.

```{r}
harm <- merge(fatal, injury, by = "EVTYPE")
as.character(harm$EVTYPE)
harm2 <- harm[order(harm$FATALITIES, decreasing = TRUE), ]
harm2
```

The below code modifys and combines the injury and fatal data frames, keeping track of which came from which.  They are then combined into one data frame, plot_data (shown below), and that is plotted in a bar chart.  

```{r}
type <- c("INJURIES", "INJURIES", "INJURIES", "INJURIES", "INJURIES")
injury1 <- cbind(injury, type)
type <- c("FATALITIES", "FATALITIES", "FATALITIES", "FATALITIES", "FATALITIES")
fatal1 <- cbind(fatal, type)
colnames(injury1) <- c("Event", "number", "type")
colnames(fatal1) <- c("Event", "number", "type")
plot_data <- rbind(injury1, fatal1)
plot_data
```

There are a couple columns that only have one color.  This means that while the event had enough of either injuries or fatalities to show up for one, they didn't have enough to show up for the other.  Looking at the chart below it is very clear that Tornados have both the largest number of fatalities and injuries by a landslide.

```{r}
qplot(x=Event, y=number, fill=type,
      data=plot_data, geom="bar", stat="identity",
      position="dodge", ylab="Number of People", xlab = "Event")
```

### Question 2

Across the United States, which types of events have the greatest economic consequences?

To answer this question, first identify the relevent variables, in this case they are Property Damage (PROPDMG) and Crop Damage (CROPDMG).  When looking at these variables, it is apparent that you need to use the columns next to them (PROPDMGEXP and CROPDMGEXP) to figure out the correct ammount.  The values in those two columns are either a K, M or B meaning that the damage is either in the thousands, millions, or billions.  The below code takes this into account, and now all the values are ready to be compared together.  I have created a copy of my data (as StormData1) so that if anything happened, I would still have access to the original data.

```{r}
StormData1 <- StormData

StormData1[StormData1$PROPDMGEXP == "K",]$PROPDMG <- StormData1[StormData1$PROPDMGEXP == "K",]$PROPDMG * 1000
StormData1[StormData1$PROPDMGEXP == "M",]$PROPDMG <- StormData1[StormData1$PROPDMGEXP == "M",]$PROPDMG * 1000000
StormData1[StormData1$PROPDMGEXP == "B",]$PROPDMG <- StormData1[StormData1$PROPDMGEXP == "B",]$PROPDMG * 1000000000
StormData1[StormData1$CROPDMGEXP == "K",]$CROPDMG <- StormData1[StormData1$CROPDMGEXP == "K",]$CROPDMG * 1000
StormData1[StormData1$CROPDMGEXP == "M",]$CROPDMG <- StormData1[StormData1$CROPDMGEXP == "M",]$CROPDMG * 1000000
StormData1[StormData1$CROPDMGEXP == "B",]$CROPDMG <- StormData1[StormData1$CROPDMGEXP == "B",]$CROPDMG * 1000000000
```

Next using the function aggregate, the cost of both property damage and crop damage are combined for each event.  

```{r}
Cost <- aggregate(CROPDMG + PROPDMG ~ EVTYPE, StormData1, sum)
cost <- subset(Cost, Cost[, 2] > 10000000000)
colnames(cost) <- c("Event", "cost")
cost
```

The plot below shows the calculated cost for the 9 events that cost the most.

```{r}
qplot(x=Event, y=cost, data=cost, geom="bar", stat="identity", 
      position="dodge", main="Total Property & Crop Damage", 
      ylab = "Cost in Dollars", xlab = "Event")
```

### Results

Using this data, the most harmful event to population health is Tornados.  They cause more injuries and fatalities than any other desaster.  The next closes would possibly be Thunderstorm Wind (TSTM WIND) and Excessive Heat (EXCESSIVE HEAT).

The event that has the largest economic impact is Flooding followed by Hurricane/Typhoon and Tornado.

